---
title: "CSP-571 Project Spring 24"
author: "Kyung Jin Kwak (kkwak4@hawk.iit.edu), Srujan Ramesh (sramesh19@hawk.iit.edu)"
output: 
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
---

## Install necessary libraries

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(formattable)
library(randomForest)
library(leaps)
```

## Loading & Checking Data

```{r}
taxi_df <- read.csv("../taxi_Trips__2024.csv")
```

```{r}
head(taxi_df)
```

`{r}tlsg print(names(taxi_df)) print(nrow(taxi_df))`

```{r}
colSums(is.na(taxi_df))
```

```{r}
(unique(taxi_df$Company))
```

## Understanding the features

```{r}
feature_desc <- read.csv("taxi_Trips_2024_Feature_descriptions.csv")
feature_desc
```

```{r}
notes <- list()
for (feature in names(taxi_df)) {
  
  curr_note <- paste("Valid rows:",(nrow(taxi_df) - sum(is.na(taxi_df[,feature]))),
                     "; N/A rows:", sum(is.na(taxi_df[,feature])),
                     "; Unique values:", length(unique(taxi_df[,feature])))
  notes <- append(notes, curr_note)
}
notes
```

```{r}
feature_desc$Notes <- unlist(notes)
feature_desc
```

```{r}
# Install the formattable package if not already installed
if (!require(formattable)) {
  install.packages("formattable")
}

formattable(feature_desc, align = c("l", "l", "l", "l"), list(Notes = formatter("span", style = "color:blue"), Type = formatter("span", style = "color:green"))
)

```

```{r}
length(which(taxi_df$Fare + taxi_df$Tips + taxi_df$Tolls + taxi_df$Extras != taxi_df$Trip.Total))
```

## 1. Data Cleaning

### 1.1 Dropping unwanted features

```{r}
features_to_drop <- c("Trip.End.Timestamp", "Pickup.Census.Tract", "Dropoff.Census.Tract", "Pickup.Centroid.Latitude", "Pickup.Centroid.Longitude", "Pickup.Centroid.Location", "Dropoff.Centroid.Latitude", "Dropoff.Centroid.Longitude", "Dropoff.Centroid..Location")

simplified_taxi_df <- subset(taxi_df, select = -c(Trip.End.Timestamp, Pickup.Census.Tract, Dropoff.Census.Tract, Pickup.Centroid.Latitude, Pickup.Centroid.Longitude, Pickup.Centroid.Location, Dropoff.Centroid.Latitude, Dropoff.Centroid.Longitude, Dropoff.Centroid..Location))
head(simplified_taxi_df)
```

### 1.2 Removing Null rows

```{r}
dim(simplified_taxi_df)
colSums(is.na(simplified_taxi_df))
```

```{r}
cleaned_taxi_df <- simplified_taxi_df[!apply(is.na(simplified_taxi_df), 1, any), ]
colSums(is.na(cleaned_taxi_df))
dim(cleaned_taxi_df)
head(cleaned_taxi_df)
```

### 1.3 Converting seconds to mins

```{r}
cleaned_taxi_df$Trip.Minutes <- round(cleaned_taxi_df$Trip.Seconds / 60, digits = 2) 
cleaned_taxi_df$Trip.Seconds <- NULL 
head(cleaned_taxi_df)
```

### 1.4 Converting timestamp to hour of the day & day of the week

```{r}
cleaned_taxi_df$Trip.Start.Timestamp <- as.POSIXct(cleaned_taxi_df$Trip.Start.Timestamp, 
                                                    format = "%m/%d/%Y %I:%M:%S %p")

# ---------------------------------------------------------------------------------
# If you are facing any errors with the format used above, try using the one below 
# ---------------------------------------------------------------------------------
# cleaned_taxi_df$Trip.Start.Timestamp <- as.POSIXct(cleaned_taxi_df$Trip.Start.Timestamp, format = "%m/%d/%y %H:%M")


cleaned_taxi_df$Trip.Start.Date <- as.Date(cleaned_taxi_df$Trip.Start.Timestamp)

cleaned_taxi_df$Trip.Hour.Of.The.Day <- as.integer(format(cleaned_taxi_df$Trip.Start.Timestamp, format = "%H"))
cleaned_taxi_df$Trip.Hour.Of.The.Day <- as.factor(cleaned_taxi_df$Trip.Hour.Of.The.Day)

days_of_week <- c("Sunday" = 1, "Monday" = 2, "Tuesday" = 3, "Wednesday" = 4, "Thursday" = 5, "Friday" = 6, "Saturday" = 7)
cleaned_taxi_df$Trip.Day.Of.The.Week <- as.integer(days_of_week[weekdays(cleaned_taxi_df$Trip.Start.Timestamp)])
cleaned_taxi_df$Trip.Day.Of.The.Week <- as.factor(cleaned_taxi_df$Trip.Day.Of.The.Week)
head(cleaned_taxi_df)
```

### 1.5 Convert datatype of payment type

```{r}
cleaned_taxi_df$Payment.Type <- as.factor(cleaned_taxi_df$Payment.Type)
cleaned_taxi_df$Company <- as.factor(cleaned_taxi_df$Company)
cleaned_taxi_df$Taxi.ID <- as.factor(cleaned_taxi_df$Taxi.ID)
head(cleaned_taxi_df)
```

### 1.6 Convert datatype of community area to string

```{r}
cleaned_taxi_df$Pickup.Community.Area <- as.factor(cleaned_taxi_df$Pickup.Community.Area)
cleaned_taxi_df$Dropoff.Community.Area <- as.factor(cleaned_taxi_df$Dropoff.Community.Area)
```

### 1.7 Check dimension and summary of the cleaned dataset

```{r}
dim(cleaned_taxi_df)
names(cleaned_taxi_df)
summary(cleaned_taxi_df)
head(cleaned_taxi_df)
```

```{r}
saveRDS(cleaned_taxi_df, "taxi_df_data_cleaning.rds")
```


## 2. Exploratory Data Analysis

### 2.1 Check and remove rows that has outliers

```{r}
cleaned_taxi_df <- readRDS("taxi_df_data_cleaning.rds")
attach(cleaned_taxi_df)
```

```{r}
hist(Trip.Total, breaks = 50, main = "Histogram of Total fare", xlab = "Total Fare")
summary(Trip.Total)
```

Looking at the histogram and the boxplot, there are some extreme values that hinders accuracy of our future model, hence needs to be removed.

```{r}
fare_related_features <- cleaned_taxi_df[, c('Fare','Tips','Tolls','Extras','Trip.Total')]
boxplot(fare_related_features)
```

```{r}
# Define the function to detect outliers for a single column
is.outlier <- function(x) {
  iqr <- IQR(x, na.rm = TRUE)
  lower <- quantile(x, 0.25, na.rm = TRUE) - 1.5 * iqr
  upper <- quantile(x, 0.75, na.rm = TRUE) + 1.5 * iqr
  return(x < lower | x > upper)
}

outliers <- is.outlier(Trip.Total)
```

number of outliers

```{r}
sum(outliers)
```

Remove all the rows that are outliers of Trip.Total

```{r}
cleaned_taxi_df <- cleaned_taxi_df[!outliers, ]
dim(cleaned_taxi_df)
```

### 2.2 Histrogram of each features

```{r}
hist(cleaned_taxi_df$Fare, breaks = 25, main = "Histogram of Fare", xlab = "Fare")
summary(cleaned_taxi_df$Fare)

hist(cleaned_taxi_df$Tips, breaks = 50, main = "Histogram of Tips", xlab = "Tips")
summary(cleaned_taxi_df$Tips)

hist(cleaned_taxi_df$Tolls, breaks = 50, main = "Histogram of Tolls", xlab = "Tolls")
summary(cleaned_taxi_df$Tolls)

hist(cleaned_taxi_df$Extras, breaks = 50, main = "Histogram of Extras", xlab = "Extras")
summary(cleaned_taxi_df$Extras)

hist(cleaned_taxi_df$Trip.Total, breaks = 25, main = "Histogram of Trip.Total", xlab = "Trip.Total")
summary(cleaned_taxi_df$Trip.Total)
```

```{r}
boxplot(cleaned_taxi_df[, c('Fare','Tips','Tolls','Extras','Trip.Total')])
```

### 2.3 Heatmap of correlation of each features

```{r}
# Select only numeric columns for correlation
numeric_columns <- sapply(cleaned_taxi_df, is.numeric)
cor_matrix <- cor(cleaned_taxi_df[, numeric_columns], use = "complete.obs")

print(cor_matrix)

# Convert the correlation matrix to a long format for ggplot2
cor_data <- as.data.frame(as.table(cor_matrix))

# Create a heatmap with ggplot2
ggplot(cor_data, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1,1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Correlation Matrix Heatmap")
```

Trip.Total is in high correlation relationship with 'Fare, Trip.Miles, Tips', but almost no relationship with 'Tolls, Hour of the trip'

### 2.4 Other data exploration

#### 1) Average Number of Trips per Taxi in a Day Over Time

```{r}

# Generate a sequence of dates within the range of your data
date_range <- seq(min(cleaned_taxi_df$Trip.Start.Date), max(cleaned_taxi_df$Trip.Start.Date), by = "day")

# Create a data frame of weekend dates
weekend_dates <- data.frame(Date = date_range[weekdays(date_range) %in% c("Saturday", "Sunday")])

# Group by Taxi ID and Date, then summarize the average trips
average_trips_per_taxi <- cleaned_taxi_df %>%
  group_by(Taxi.ID, Trip.Start.Date) %>%
  summarise(Trips = n(), .groups = 'drop') %>%
  group_by(Trip.Start.Date) %>%
  summarise(AvgTrips = mean(Trips), .groups = 'drop')

# Plot the average trips per taxi over time
plot <- ggplot(average_trips_per_taxi, aes(x = Trip.Start.Date, y = AvgTrips)) +
  geom_line() +
  labs(title = "Average Number of Trips per Taxi in a Day Over Time",
       x = "Date",
       y = "Average Trips per Taxi") +
  theme_minimal()

# Add vertical lines for weekends
plot + geom_vline(data = weekend_dates, aes(xintercept = as.numeric(Date)), color = "red", linetype = "dashed")
 
```

#### 2) Total Trips Over Time

```{r}
# First, calculate the total trips per day
total_trips_over_time <- cleaned_taxi_df %>%
  group_by(Trip.Start.Date) %>%
  summarise(TotalTrips = n(), .groups = 'drop')

# Plot the total trips over time
total_trips_plot <- ggplot(total_trips_over_time, aes(x = Trip.Start.Date, y = TotalTrips)) +
  geom_line() +
  labs(title = "Total Trips Over Time",
       x = "Date",
       y = "Total Trips") +
  theme_minimal()

# Add vertical lines for weekends
total_trips_plot + geom_vline(data = weekend_dates, aes(xintercept = as.numeric(Date)), color = "red", linetype = "dashed")
```

We can see that people tends not to use taxi on weekends than weekdays.

#### 3) Pickup and Dropoff Area Count - Overlaid Histogram

```{r fig.width=14, fig.height=7}
# Create a long format data frame for pickup and dropoff areas
area_data <- tidyr::pivot_longer(
  cleaned_taxi_df, 
  cols = c("Pickup.Community.Area", "Dropoff.Community.Area"), 
  names_to = "AreaType", 
  values_to = "Area"
)

# Plot overlaid bar charts for pickup and dropoff area counts
ggplot(area_data, aes(x = as.factor(Area), fill = AreaType)) +
  geom_bar(position = "identity", alpha = 0.5) +  # Set alpha for transparency
  scale_fill_manual(values = c("Pickup.Community.Area" = "blue", "Dropoff.Community.Area" = "red")) +
  labs(title = "Overlaid Bar Chart of Pickup and Dropoff Area Counts",
       x = "Area Code",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

```

#### 4) Trip.Total by Payment Type

```{r}
# Summarise Trip.Total by Payment Type
trip_total_by_PaymentType <- cleaned_taxi_df %>%
  group_by(Payment.Type) %>%
  summarise(TotalRevenue = sum(Trip.Total), .groups = 'drop')

# Create a bar plot of Trip Total by Payment Type
ggplot(trip_total_by_PaymentType, aes(x = Payment.Type, y = TotalRevenue, fill = Payment.Type)) +
  geom_col() +  # This creates a bar chart with pre-summarized data
  labs(title = "Trip Total by Payment Type",
       x = "Payment Type",
       y = "Total Revenue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

#### 5) Trip.Total by Day of the week

```{r}
trip_total_by_DayoftheWeek <- cleaned_taxi_df %>%
  group_by(Trip.Day.Of.The.Week) %>%
  summarise(TotalRevenue = sum(Trip.Total), .groups = 'drop')

# Define a named vector to map day numbers to day names
day_names <- c("1" = "Sunday", "2" = "Monday", "3" = "Tuesday", "4" = "Wednesday", 
               "5" = "Thursday", "6" = "Friday", "7" = "Saturday")

# Create the bar plot, using the named vector for axis labels
ggplot(trip_total_by_DayoftheWeek, aes(x = Trip.Day.Of.The.Week, y = TotalRevenue, fill = Trip.Day.Of.The.Week)) +
  geom_col() +
  scale_x_discrete(labels = day_names) +  # Use the day_names vector for axis labels
  labs(title = "Trip Total by Day of the Week",
       x = " ",
       y = "Total Revenue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### 6) Trip.Total by Company

```{r}
# Summarise Trip.Total by Company and arrange by TotalRevenue in descending order
trip_total_by_Company <- cleaned_taxi_df %>%
  group_by(Company) %>%
  summarise(TotalRevenue = sum(Trip.Total), .groups = 'drop') %>%
  arrange(desc(TotalRevenue))

# Display the full table sorted by Total Revenue
kable(trip_total_by_Company)
```

#### 7) Trip.Total by time in a day

```{r}
# Summarise Trip.Total by Company and arrange by TotalRevenue in descending order
trip_total_by_time <- cleaned_taxi_df %>%
  group_by(Trip.Hour.Of.The.Day) %>%
  summarise(TotalRevenue = sum(Trip.Total), .groups = 'drop') %>%
  arrange(desc(TotalRevenue))

ggplot(trip_total_by_time, aes(x = Trip.Hour.Of.The.Day, y = TotalRevenue, fill = TotalRevenue)) +
  geom_col() +
  labs(title = "Trips total by time in a day",
       x = "Time",
       y = "Trips total") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") + # This line removes the legend
  scale_fill_gradient(low = "lightblue", high = "blue") # This line adds a color gradient from light blue to blue
```

#### 7) Trips Count by Paymeny type

```{r}
# Summarise Trip id by Payment Type
trip_count_by_PaymentType <- cleaned_taxi_df %>%
  group_by(Payment.Type) %>%
  summarise(TripCount = n_distinct(Trip.ID))

# Create a bar plot of Trip Total by Payment Type
ggplot(trip_count_by_PaymentType, aes(x = Payment.Type, y = TripCount, fill = Payment.Type)) +
  geom_col() +  # This creates a bar chart with pre-summarized data
  labs(title = "Trips Count by Paymeny type",
       x = "Payment Type",
       y = "Trip Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

#### 8) Trips Count by day of the week

```{r}
trip_count_by_DayoftheWeek <- cleaned_taxi_df %>%
  group_by(Trip.Day.Of.The.Week) %>%
  summarise(TripCount = n_distinct(Trip.ID))

# Define a named vector to map day numbers to day names
day_names <- c("1" = "Sunday", "2" = "Monday", "3" = "Tuesday", "4" = "Wednesday", 
               "5" = "Thursday", "6" = "Friday", "7" = "Saturday")

# Create the bar plot, using the named vector for axis labels
ggplot(trip_count_by_DayoftheWeek, aes(x = Trip.Day.Of.The.Week, y = TripCount, fill = Trip.Day.Of.The.Week)) +
  geom_col() +
  scale_x_discrete(labels = day_names) +  # Use the day_names vector for axis labels
  labs(title = "Trip Count by Day of the Week",
       x = " ",
       y = "Trip Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### 9) Trips Count by Company

```{r}
# Summarise Trip.Total by Company and arrange by TotalRevenue in descending order
trip_count_by_Company <- cleaned_taxi_df %>%
  group_by(Company) %>%
  summarise(TripCount = n_distinct(Trip.ID), .groups = 'drop') %>%
  arrange(desc(TripCount))

# Display the full table sorted by Total Revenue
kable(trip_count_by_Company)
```

#### 10) Trips Count by time in a day

```{r}
# Summarise Trip.Total by Company and arrange by TotalRevenue in descending order
trip_count_by_time <- cleaned_taxi_df %>%
  group_by(Trip.Hour.Of.The.Day) %>%
  summarise(TripCount = n_distinct(Trip.ID), .groups = 'drop') %>%
  arrange(desc(TripCount))

ggplot(trip_count_by_time, aes(x = Trip.Hour.Of.The.Day, y = TripCount, fill = TripCount)) +
  geom_col() +
  labs(title = "Trips Count by time in a day",
       x = "Time",
       y = "Trip Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") + # This line removes the legend
  scale_fill_gradient(low = "lightblue", high = "blue") # This line adds a color gradient from light blue to blue
```

```{r}
detach(cleaned_taxi_df)
saveRDS(cleaned_taxi_df, "taxi_df_EDA.rds")
```


Some Key Findings from EDA

1.  Total taxi fare correlates with original fare, trip miles, tips, where does not correlates with tolls and trip hours of the day.

2.  People tends to use taxi on weekdays (especially on Wed \> Tue \> Mon), but not on weenkends.

3.  Majority of people use Credit Card then Cash to pay the taxi fare.

4.  The most popular Pickup area are '8, 76, 32, 28'.

5.  The most popular Dropoff area are '8, 32, 28'.

6.  Top 5 most frequently used taxi companies in Chicago are Flash Cab, Taxi Affiliation Services, Sun Taxi, Taxicab Insurance Agency Llc, and City Service.

7.  The peak time for using taxi is 4pm\~5pm. But by the taxi fare, the peak time is 3pm\~7pm.

## 3. Model Selection

### 3.1 Feature Selection using Random Forests

```{r}
cleaned_taxi_df <- readRDS("taxi_df_EDA.rds")
```

```{r}
features_to_drop <- c('Trip.ID', 'Taxi.ID', 'Trip.Start.Timestamp', 'Trip.Start.Date', 'Payment.Type', 'Company', 'Tips', 'Tolls', 'Extras', 'Trip.Total')
selected_taxi_df <- cleaned_taxi_df %>% select(-one_of(features_to_drop))

factor_columns <- sapply(selected_taxi_df, is.factor)
selected_taxi_df[factor_columns] <- sapply(selected_taxi_df[factor_columns], function(x) as.numeric(levels(x))[x])

head(selected_taxi_df)
```

```{r}
set.seed(123)

sample_size <- floor(0.5 * nrow(selected_taxi_df))
sample_indices <- sample(seq_len(nrow(selected_taxi_df)), size=sample_size)

random_forest <- randomForest(Fare ~ ., data = selected_taxi_df[sample_indices, ], importance = TRUE, ntree = 25)
importance(random_forest)
```

### 3.2 Feature Selection using Best Subset Selection

```{r}
regfit.full <- regsubsets(Fare ~ ., data = selected_taxi_df)
summary(regfit.full)
```


### 3. Fare prediction using Splines

```{r}

```

### 3. Fare prediction using SVM

```{r}

```

### 3. Tip chances using Distance Trees

```{r}

```